---
title: "Exploring the abundance table"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,   # hide package/startup messages
  warning = FALSE,
  fig.width = 10,    # default plot size (inches)
  fig.height = 4.5
)
library(tidyverse)
```

# Introduction

This report explores the abundance table `table.tsv`:

- **rows** = taxa, labeled with their full taxonomic lineage
  (e.g. `k__Bacteria,p__Firmicutes,...,g__Lactobacillus`)
- **columns** = samples (sequencing runs named `ERR...`)
- **values** = raw abundance counts

We will import the data, reshape it to tidy format, parse the taxonomy,
look at library sizes and community composition, and finish with a few plots.

# Import the data

`read_tsv()` parses the tab-separated file. The first column is called
`#Taxon` in the header, so we rename it to `Taxon` right away
(backticks are needed because the name starts with `#`).

```{r import}
counts <- read_tsv("table.tsv") |>
  rename(Taxon = `#Taxon`)

cat("Table dimensions:", nrow(counts), "taxa x", ncol(counts) - 1, "samples\n")
counts |> slice(1:5, 1:5)
```

# Tidy the data

For most analyses (and for ggplot2) it is convenient to have **one row per
(taxon, sample) combination**: this is the *long* or *tidy* format.
`pivot_longer()` stacks all sample columns into two new columns: `Sample`
(the former column name) and `Count` (the value).

```{r tidy}
tidy_counts <- counts |>
  pivot_longer(
    cols = -Taxon,
    names_to = "Sample",
    values_to = "Count"
  )

tidy_counts
```

The wide table of `r nrow(counts)` taxa x `r ncol(counts) - 1` samples becomes
`r nrow(tidy_counts)` rows, of which `r sum(tidy_counts$Count > 0)` (~68%) are
nonzero.

## Parse the taxonomy

Row labels are full lineage strings. We extract the rank prefixes
(`p__` = phylum, `g__` = genus, ...) with regular expressions:

- `str_extract(Taxon, "g__[^,]+")` finds the `g__Genus` token
  (`[^,]+` = any characters up to the next comma)
- `str_remove(..., "g__")` strips the prefix
- `replace_na(...)` labels rows that lack the rank (e.g. `u__unclassified`)

```{r taxonomy}
tidy_counts <- tidy_counts |>
  mutate(
    Phylum = str_extract(Taxon, "p__[^,]+") |> str_remove("p__") |>
      replace_na("unclassified"),
    Genus  = str_extract(Taxon, "g__[^,]+") |> str_remove("g__") |>
      replace_na("unclassified")
  )

tidy_counts
```

# Exploration

## Library sizes

The **library size** is the total number of counts of a sample. Samples can
differ enormously in sequencing depth, which matters when comparing them ---
hence the relative abundances used later.

```{r library-sizes}
library_sizes <- tidy_counts |>
  summarise(Total = sum(Count), .by = Sample) |>
  arrange(desc(Total))

library_sizes
```

The largest sample (`r library_sizes$Sample[1]`,
`r format(library_sizes$Total[1], big.mark = ",")` counts) has more than
**100x** the reads of the smallest one
(`r format(min(library_sizes$Total), big.mark = ",")` counts).

## Composition by phylum

Collapsing all taxa that share the same phylum gives the broadest picture of
the community:

```{r phylum}
by_phylum <- tidy_counts |>
  summarise(Total = sum(Count), .by = Phylum) |>
  arrange(desc(Total)) |>
  mutate(Percent = round(100 * Total / sum(Total), 2))

by_phylum
```

The community is dominated by **Firmicutes** (~89%), followed by
*Proteobacteria* (~9%): a composition typical of fermented-food /
dairy-type microbiomes.

## Top genera

```{r top-genera}
top_genera <- tidy_counts |>
  summarise(Total = sum(Count), .by = Genus) |>
  arrange(desc(Total)) |>
  slice_head(n = 10)

top_genera
```

*Lactobacillus* alone accounts for more than half of all counts.

# Plots

## Library size per sample

Totals span more than two orders of magnitude, so the y-axis uses a
**log scale**: otherwise the shallow samples would be invisible.

```{r plot-library, fig.height=4}
ggplot(library_sizes, aes(x = reorder(Sample, Total), y = Total)) +
  geom_col(fill = "steelblue") +
  scale_y_log10() +
  labs(
    title = "Library size per sample",
    x = NULL,
    y = "Total counts (log10 scale)"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6))
```

## Top 10 genera by total abundance

`reorder()` sorts the bars by abundance, and `coord_flip()` makes them
horizontal so the genus names are easy to read.

```{r plot-genera, fig.height=5, fig.width=7}
ggplot(top_genera, aes(x = reorder(Genus, Total), y = Total)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Top 10 genera by total abundance",
    x = NULL,
    y = "Total counts (all samples)"
  ) +
  theme_minimal(base_size = 11)
```

## Relative abundance per sample

Converting counts to **percentages** removes the effect of unequal sequencing
depth and makes community composition comparable across samples. The stacked
bars show the top 8 genera; the composition differences between samples are
striking: some are almost pure *Lactobacillus*, others are dominated by
*Lactococcus* or by *Leuconostoc*/*Weissella*.

```{r plot-composition, fig.height=5}
top8 <- top_genera$Genus[1:8]

composition <- tidy_counts |>
  mutate(RelAbundance = 100 * Count / sum(Count), .by = Sample) |>
  filter(Genus %in% top8) |>
  # Sort genera in the legend by overall abundance
  mutate(Genus = factor(Genus, levels = rev(top8)))

ggplot(composition, aes(x = Sample, y = RelAbundance, fill = Genus)) +
  geom_col(width = 0.9) +
  labs(
    title = "Top 8 genera: relative abundance per sample",
    x = NULL,
    y = "Relative abundance (%)",
    fill = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6),
    legend.position = "right"
  )
```

# Recap

| Step | Key functions |
|---|---|
| Import | `read_tsv()`, `rename()` |
| Reshape | `pivot_longer()` |
| Parse taxonomy | `str_extract()`, `str_remove()`, `replace_na()` |
| Summarize | `summarise(.by = ...)`, `arrange()`, `slice_head()` |
| Transform | `mutate()` (incl. grouped with `.by`), `filter()` |
| Plot | `ggplot()` + `geom_col()`, `scale_y_log10()`, `coord_flip()` |
